Power and sample size for longitudinal data: the ``longpower" R package and app

### 2020-10-29 Manuscript/Manuscript-Draft.tex
update manuscript
some edits to draft manuscript

	% Title of paper
	\title{A tutorial on the ``longpower" R package and a web-based App for conducting power analysis and sample size computation for longitudinal data}

@@

	% Running headers of paper:
	\markboth%
	% First field is the short list of authors
	{S. Iddi and others}
	% Second field is the short title of the paper
	{A tutorial on the ``longpower" package and web-based App}

@@

	\begin{abstract}
		{Longitudinal randomized trials are common in medical and clinical research. These trials often require the collection of data on the same individual at different data collection waves. At the design phase of such studies, sample size computations are critical to ensure that studies are sufficiently powered to provide reliable and valid inference. There are several methodologies for calculating sample sizes for longitudinal studies that depend on many considerations including the type of research design, outcome type, and proposed analytical methods. More advanced procedures take into account several factors leading to more complicated formulas for computing sample sizes. To provide easy access to common and widely used sample size and power calculation formulas, this tutorial briefly describes some methods for longitudinal data.  We also enrich the discussion with real-life examples comparing treatment versus control groups in randomized trials assessing treatment effect on clinical outcomes. Accompanying this tutorial is also a web-based sample size shinyApp developed to help researchers to conduct different sample size and power calculations by allowing user-specified parameters or pilot parameters generated from the Alzheimer’s Disease Neuroimaging Initiative (ADNI) study.}
		{Mixed model for repeated measures, linear mixed model, shinyApp, power, sample size, longpower, longitudinal data}
	\end{abstract}

\section{Introduction}
\label{sec1}
Longitudinal designs are generally preferred over cross-sectional research design as they yield higher statistical power. As such, many biomedical and medical studies employ longitudinal design to study changes over time in outcome at the individual, group, or population level. In the early stages in the design of a longitudinal experimental or population study, it is imperative to ensure that sufficient participants are recruited to detect a predetermined effect size.  Inadequate sample sizes lead to invalid or inconclusive inference and a complete waste of resources (\cite{Lu_Methrotra_Liu(2009),Yan_Su(2006)}). On the other hand,  oversampling causing a waste of resources and exposure to many participants to harm that may be associated with the research (\cite{Lu_Methrotra_Liu(2009)}). Thus, optimal sample size and power analysis have become an important prerequisite for any quantitative research design. Not only are these required during the design phase of research, but it has also become mandatory when preparing protocols for ethical review and research grant applications to guarantee both economic and ethical benefits. 
Determining the right sample size for a study is not a straightforward task. Despite the plethora of sample size formulas for repeated measures (\cite{Overall_Doyle(1994),Lui(1992),Rochon(1991), Guo_etal(2013)}), cluster repeated measures (\cite{Liu_Shih_Gehan(2002)}), multivariate repeated measures (\cite{Vonesh_Schork(1986),Guo_Johnson(1996)}), longitudinal research designs (\cite{Lefante(1990)}), the tasks of gathering the necessary inputs and getting the right software to carry out the computation are challenging to many. Investment in such efforts is not what many non-technical researchers desire to embark on. Commonly, the easiest route taken by most research is to use sample size formulas for very basic cross-sectional studies and adjust for design effect due to repeated measures. Although such approaches are valid, the best approach is the use of formulas derived directly from models for longitudinal or repeated measures to align with planned data analysis and yielding greater statistical power. 
Several factors need to be considered before choosing the right sample size formulas that increase the statistical power of a study. \cite{Guo_etal(2013)} describe practical methods for the selection of appropriate sample size for repeated measures addressing issues of missing data, and the inclusion of more than one covariates to control for differences in response at baseline.

@@

\subsection{A Web-based App}
\label{model4}
This ShinyApp dashboard is developed to easily generate sample size and conduct power analysis for a longitudinal study design with two-group comparisons for a continuous outcome. The App implements the sample size formula of \cite{Liu_Liang(1997)} and \cite{Diggle_etal(1994)} using functions developed for the R `longpower' package. The `longpower' package handles cases where time is treated either as continuous or categorical. The former approach uses the linear mixed model with random intercept and slope while the later leads to the well-known Mixed Model of Repeated Measures (MMRM) used in many clinical trial applications for conducting statistical analysis.

>>>>

	% Title of paper
	\title{Power and sample size for longitudinal data: the ``longpower" R package and app}

@@

	% Running headers of paper:
	\markboth%
	% First field is the short list of authors
	{S. Iddi and others}
	% Second field is the short title of the paper
	{Power and sample size for longitudinal data}

@@

	\begin{abstract}
		{Longitudinal studies are ubiquitous in medical and clinical research. Sample size computations are critical to ensure that these studies are sufficiently powered to provide reliable and valid inference. There are several methodologies for calculating sample sizes for longitudinal studies that depend on many considerations including the type of research design, outcome type, and proposed analytical methods. More advanced procedures take into account several factors leading to more complicated formulas for computing sample sizes. To provide easy access to common and widely used sample size and power calculation formulas, this tutorial briefly describes some methods for longitudinal data.  We also enrich the discussion with real-life examples comparing treatment versus control groups in randomized trials assessing treatment effect on clinical outcomes. Accompanying this tutorial is also a web-based sample size Shiny app developed to help researchers to conduct different sample size and power calculations by allowing user-specified parameters or pilot parameter estimation using data from the Alzheimer’s Disease Neuroimaging Initiative (ADNI) study.}
		{Mixed model for repeated measures, linear mixed model, Shiny app, power, sample size, longpower, longitudinal data}
	\end{abstract}

\section{Introduction}
\label{sec1}
Longitudinal designs are generally preferred over cross-sectional research design as they yield provide richer data and greater statistical power. As such, many biomedical and medical studies employ longitudinal design to study changes over time in outcome at the individual, group, or population level. Early in the design of a longitudinal experimental or natural history study, it is imperative to ensure that the study is adequately powered for its aims.  Inadequate sample sizes leads to invalid or inconclusive inference and squandered resources (\cite{Lu_Methrotra_Liu(2009),Yan_Su(2006)}). On the other hand,  oversampling causing a waste of resources and exposure to many participants to harm that may be associated with the research (\cite{Lu_Methrotra_Liu(2009)}). Thus, optimal sample size and power analysis have become an important prerequisite for any quantitative research design. Not only are these required during the design phase of research, but it has also become mandatory when preparing protocols for ethical review and research grant applications to guarantee both economic and ethical benefits. 
Determining the right sample size for a study is not a straightforward task. Despite the plethora of sample size formulas for repeated measures (\cite{Overall_Doyle(1994),Lui(1992),Rochon(1991), Guo_etal(2013)}), cluster repeated measures (\cite{Liu_Shih_Gehan(2002)}), multivariate repeated measures (\cite{Vonesh_Schork(1986),Guo_Johnson(1996)}), longitudinal research designs (\cite{Lefante(1990)}), the tasks of gathering the necessary inputs and getting the right software to carry out the computation are challenging to many. Investment in such efforts is not what many non-technical researchers desire to embark on. Commonly, the easiest route taken by most research is to use sample size formulas for very basic cross-sectional studies and adjust for design effect due to repeated measures. Although such approaches are valid, the best approach is the use of formulas derived directly from models for longitudinal or repeated measures to align with planned data analysis and yielding greater statistical power. 
Several factors need to be considered before choosing the right sample size formulas that increase the statistical power of a study. \cite{Guo_etal(2013)} describe practical methods for the selection of appropriate sample size for repeated measures addressing issues of missing data, and the inclusion of more than one covariates to control for differences in response at baseline.

@@

\subsection{A Web-based App}
\label{model4}
This Shiny app dashboard is developed to easily generate sample size and conduct power analysis for a longitudinal study design with two-group comparisons for a continuous outcome. The App implements the sample size formula of \cite{Liu_Liang(1997)} and \cite{Diggle_etal(1994)} using functions developed for the R `longpower' package. The `longpower' package handles cases where time is treated either as continuous or categorical. The former approach uses the linear mixed model with random intercept and slope while the later leads to the well-known Mixed Model of Repeated Measures (MMRM) used in many clinical trial applications for conducting statistical analysis.
